---
title: "HOMEWORK 2"
format: pdf
editor: visual
---

```{r warning=FALSE}
library(tidyverse)
library(igraph)
library(igraphdata)
library(jsonlite)
library(visNetwork)
library(widgetframe)
library(ggthemes)
library(readr)
```

# IceLab, Department of Physics, Umea University

# Description: Network data obtained from a web community about sexual activity

# Completely bi-partite network with multiple edges

# Nodes: female (sex-sellers) and male (sex-buyers)

# Edges: if a male writes a post about a sexual encounter with a female

# 

# Originally published in:

# "Information dynamics shape the sexual networks of Internet-mediated prostitution"

# L E C Rocha, F Liljeros, and P Holme

# Proceedings of the National Academy of Sciences of the USA 107 (13), 5706-5711 (2010)

```{r}
raw <- read_delim(
  file       = "brazilian_prostitution.csv",      
  delim      = ";",
  comment    = "#",
  col_names  = FALSE,
  trim_ws    = TRUE,
  col_types  = cols(.default = col_integer())
) |> 
    rename(
    `Female-id`                                                                       = X1,
    `Male-id`                                                                         = X2,
    `Date (in days) of posting`                                                       = X3,
    `Female grade given by the male`                  = X4,
    `Anal sex with/without condom?`  = X5,
    `Oral sex with condom?`          = X6,
    `Mouth kiss?`                    = X7
  ) |> 
  select(-`Date (in days) of posting`)


raw

```

```{r}
g <- graph_from_data_frame(raw, directed=FALSE)
```

```{r}
visIgraph(g, physics = F) |> 
visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE)
```
