---
title: "Social Network Analysis | UC3M"
subtitle: "Final Project: Spreading on a Social Network"
author: 
      - Samantha Riordan   | 100572578 
      - Miriam Lim-Leal    | 100453357
      - Noah Emanuel-Fürup | 100564888
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output: 
  html_document: 
  toc: TRUE
  toc_float: TRUE
  toc_depth: 3
  number_sections: TRUE
---

## Objective: Study the problem of spreading on a social network

## Intro

```{r, include=F}
# Libraries 
library(tidyverse)
library(igraph)
library(ggthemes)
library(moments)
library(knitr)
```

**Network**: Social Circles in Ego Networks on Facebook.

McAuley, J. & Leskovec, J. (2012). Learning to discover social circles in ego networks. *Advances in Neural Information Processing Systems, 25*, 539-547. <https://doi.org/10.48550/arXiv.1210.8182>

### Network Setup

```{r}
# Load Network 
edgelist <- read.table("facebook_combined.txt")
# Create the graph 
g <- graph_from_data_frame(edgelist, directed = FALSE)
g <- simplify(g)

# Check number of nodes and edges
vcount(g) # Number of nodes 
ecount(g) # Number of edges

# Check if the network connected
is_connected(g) #TRUE

deg <- degree(g, mode = "all")
mean(deg)
sd(deg)
median(deg)
# Check skew
skewness(deg) 
```

### Network Attributes

The network returns a total of 4,039 nodes and 88,234 links. The total number of nodes represent individual Facebook profiles and the edges are an undirected representation of Facebook friendship connections between profiles. The mean degree (M = 43.69, 2dp), median (Mdn = 25) and standard deviation (sd = 52.42, 2dp) indicated a positively skewed degree distribution with high variability.

A skewness coefficient of 4.52 (2dp) confirmed the positive skew and is consistent with the presence of highly concentrated degrees typical of social networks. To acknowledge this, the epidemic threshold is computed using the heterogeneous mean-field formula β_c = μ⟨k⟩/(⟨k²⟩ − ⟨k⟩), which accounts for degree variance. Furthermore, the evidence of the skew further motivates the use of degree-based centrality for targeted seeding and immunisation strategies in subsequent analyses, where highly central nodes are expected to play a disproportionate role in spreading dynamics.

### Interpreting the Network 

The analysis for this task interprets misinformation spread as analogous to an infectious process. Specifically, susceptible (S) nodes represent individuals not yet exposed to the misinformation; infected (I) nodes represent active spreaders; and recovered (R) nodes represent individuals who have ceased spreading, via having acquired a form of digital literacy . Separately, immunisation strategies (Steps 4–5) are modelled as the deliberate removal of nodes, representing targeted interventions to prevent spread. Whilst this analogy simplifies the complexity of misinformation spread and dynamics, it provides an empirically grounded framework for the analyses of this task.

## Steps

### Step 0: SIR Function (original / no-quarantine)

```{r}
# Code from Tutorial - No change required
sim_sir <- function(g, beta, mu, seeds){
  state <- rep(0, vcount(g)) #initial state of the simulation
  state[seeds] <- 1 #infect the seeds
  t <- 0
  table <- data.frame(t = 0, inf = seeds)
  while(sum(state == 1) > 0){
    t <- t + 1
    ## I -> R
    infected <- which(state == 1) # infected nodes
    # generate a random value for every infected, if it's < mu, let the node recover.
    state[infected] <- ifelse(runif(length(infected)) < mu, 2, 1)
    
    ## S -> I
    infected <- which(state == 1) # nodes still infected
    susceptible <- which(state == 0) # susceptible nodes
    contacts <- as.numeric(unlist(adjacent_vertices(g, infected))) # contacts of infected
    contacts <- contacts[contacts %in% susceptible] # get those who are susceptible
    new_infected <- contacts[runif(length(contacts)) < beta] # infect contacts
    new_infected <- unique(new_infected) # remove repeated infections
    if(length(new_infected) > 0){
      state[new_infected] <- 1
      table <- rbind(table, 
                     data.frame(t, inf = new_infected))
    }
  }
  table
}
```

### Step 1: Epidemic Threshold

Using the heterogeneous mean-field formula the epidemic threshold was calculated as βc = 0.00095 (5 dp). Misinformation spread on social media has been estimated with Twitter specific estimates of R0 = 4.0-5.1 under the SIR model (Cinelli et al., 2020). Using the relationship β = R0 · βc, a transmission rate of β = 0.00475 corresponds to R0 = 5, placed clearly above the threshold. A value of β = 0.006 was selected for subsequent analyses to ensure β was well above the threshold but remained theoretically relevant to the specifics of the network and a socially plausible transmission parameter for modelling misinformation spread.

```{r}
# Beta_c  
mu <- 0.1
kavg <- mean(degree(g))
ksq  <- mean(degree(g)^2)
betac <- mu*kavg/(ksq - kavg)
betac

# above-threshold beta estimate 
5 * betac
```

### Step 2: Simulate SIR

Assuming that randomly-selected 1% initial spreaders, simulate the SIR model for a range below and above that threshold and plot the number of infected people as a function of β.

The plot below shows the total number of infected nodes as a function of transmission rate β, with the theoretical epidemic threshold βc = 0.00095 marked by the dashed vertical line. Below the threshold, infections remain negligible as the epidemic fails to activate throughout the network. Above the threshold, the number of infected nodes increases substantially with β, consistent with theoretical predictions of epidemic spread in heterogeneous networks. Currently, the variability observed across β values reflects the random disposition of the single realisation simulated per parameter value, where averaging over multiple realisations would produce a smoother transition and be more robust (not done due to being too computationally expensive). At the selected above-threshold value of β = 0.006, approximately 1953 nodes were infected, representing 48.4% of the network.

```{r}
# calculate for each realisation the number of total infections, trying for different values of β and fixed μ. 
set.seed(246)
# beta range
results <- map_dfr(seq(0, 0.006, 0.0001),
        \(beta){seeds <- sample(1:vcount(g),
                                vcount(g) * 0.01)
        # make a realisation for each beta
        realisation <- sim_sir(g, beta, mu, seeds) 
        # create dataframe row
        data.frame(beta, ninf = nrow(realisation)) 
          })
# Visualise
results |> 
  ggplot(aes(x = beta, y = ninf)) + 
  geom_point() +
  geom_vline(xintercept = betac, 
             linetype = 2)


# Number of infected nodes at the 0.006 above-threshold 
results_above <- results |> filter(beta == 0.006)
cat("Nodes infected:", results_above$ninf, 
    "\nPercentage of network:", 
    round(results_above$ninf / vcount(g) * 100, 1), "%")
```

### Step 3: Maximising Infection

Choose a β well-above above βc. Using centrality, communities or any other suitable metric, find a better set of 1% of seeds in the network so we get more infected people than the random case. Measure the difference of your choice with the random case as:

a)  The difference in the total number of infected people
b)  The difference in the time of the peak of infection (when most infections happen).

Given the evidence of the skew identified in the initial network attribute analysis, the degree centrality was used to select the top 1% of seeds for targeted seeding, where high-degree nodes are expected to maximise immediate spread across the network. As shown in the plot below, the targeted seeding strategy produced a sharper, earlier epidemic peak compared to random seeding. The peak infection occurred at t = 3 for targeted seeding compared to t = 11 for random seeding (difference of 8 time steps), indicating that degree-based seeds accelerate the spread of misinformation. Total infected nodes were higher for the targeted seeding (targeted = 2,270; random = 2,106; difference = 164). At β = 0.006, both strategies produced large scale epidemics, infecting approximately 50% of the network, consistent with the above-threshold transmission rate selected.

These findings support the theoretical expectation that seeding from nodes with high-degree centality is a more effective spreading strategy than random selection. The earlier peak for targeted seeding has particularly important practical implications in the context of misinformation, where high-degree spreaders saturate the network faster, leaving less time for corrective interventions before the epidemic (or in this case, misinformation spread) peaks.

```{r}
set.seed(2026)
# Seed Selection - Unchanged
# Random spreaders
seeds_random <- sample(1:vcount(g), 
                       floor(vcount(g) * 0.01)) # round down to produce whole numbers 

# centrality-based spreaders
# Change to degree if too slow
seeds_centrality <- degree(g) |> 
  enframe(name = "inf", 
          value = "degree") |> 
  slice_max(degree, 
            n = floor(vcount(g) * 0.01)) |>  # round down to produce whole numbers
  mutate(inf = match(inf, V(g)$name)) # using match as node names start at 0 

# Simulation - Raw output (to count total unique infected)
# normal simulation
raw_random <- sim_sir(g, beta = 0.006, mu = 0.1, seeds_random)
# targeted simulation
raw_targeted <- sim_sir(g, beta = 0.006, mu = 0.1, seeds_centrality$inf)

# Summarised by time (for plot and peak time)
realisation_random <- raw_random |> 
  group_by(t) |> 
  summarize(ninf = n())
realisation_targeted <- raw_targeted |> 
  group_by(t) |> 
  summarize(ninf = n())

# Plot infection curve - Unchanged
ggplot() + 
  geom_line(data = realisation_random,
            aes(x = t, y = ninf,
                col = "Random")) +
  geom_line(data = realisation_targeted,
            aes(x = t, y = ninf,
                col = "Targeted")) + 
  labs(title = "SIR Infection Curves: Random vs Targeted Seeding",
       x = "Time", 
       y = "Number Infected",
       colour = "Seeding Strategy")

# a) Difference in total infected 
total_random <- nrow(raw_random)
total_targeted <- nrow(raw_targeted)
diff_total <- total_targeted - total_random
# b) Difference in time of peak infection 
peak_random <- realisation_random$t[which.max(realisation_random$ninf)]
peak_targeted <- realisation_targeted$t[which.max(realisation_targeted$ninf)]
diff_peak <- peak_random - peak_targeted

# Display results 
data.frame(
  Metric = c("Total Infected", "Peak Infection Time"),
  Random = c(total_random, peak_random),
  Targeted = c(total_targeted, peak_targeted),
  Difference = c(diff_total, diff_peak)) |>
  kable(caption = "Comparison of Random vs Targeted Seeding")

```

### Quarantine SIR Simulation

Modification needs to:

- Modify existing SIR function to temporarily remove the nodes (acting as a quarantine) that is a random sample of susceptible nodes at \~ t = 3 (check: based on exponential growth phase - Step 3) and then reintroduce those nodes 8 steps later (e.g., \~ t = 11).
- I think we will need to select and justify what % of the susceptible population becomes quarantined.

```{r}
# Original SIR Code from Tut - Not changed 

sim_sir <- function(g, beta, mu, seeds){
  state <- rep(0, vcount(g)) #initial state of the simulation
  state[seeds] <- 1 #infect the seeds
  t <- 0
  table <- data.frame(t = 0, inf = seeds)
  while(sum(state == 1) > 0){
    t <- t + 1
    ## I -> R
    infected <- which(state == 1) # infected nodes
    # generate a random value for every infected, if it's < mu, let the node recover.
    state[infected] <- ifelse(runif(length(infected)) < mu, 2, 1)
    
    ## S -> I
    infected <- which(state == 1) # nodes still infected
    susceptible <- which(state == 0) # susceptible nodes
    contacts <- as.numeric(unlist(adjacent_vertices(g, infected))) # contacts of infected
    contacts <- contacts[contacts %in% susceptible] # get those who are susceptible
    new_infected <- contacts[runif(length(contacts)) < beta] # infect contacts
    new_infected <- unique(new_infected) # remove repeated infections
    if(length(new_infected) > 0){
      state[new_infected] <- 1
      table <- rbind(table, 
                     data.frame(t, inf = new_infected))
    }
  }
  table
}
```

### Step 4: Quarantine Strategy

Using the same β, design a “quarantine strategy”: during the exponential growth phase of the system (typically \~*t* = 3 , but check) quarantine of the susceptible population. You can model quarantine by temporally removing these nodes. Release the quarantined nodes 8 time steps later, making them susceptible again. Measure the difference with respect to no quarantine.

Workflow: Detect the exponential growth phase (using a visual inspection of the infection curve from Step 3 to confirm \~*t* = 3) and modify the Original SIR function to create the Quarantine SIR function. Repeat Step 3 for the quarantine simulation (random seeds) identifying the total infected and time of peak infection and compare this the random seed results from Step 3 (compared as with-quarantine vs. no-quarantine strategies).

```{r}
# Code from Tut - Not Changed - Modify current code to the quarantine function and "Targeted" plot to be "Quarantine" 

# normal simulation
realisation_random <- sim_sir(g, beta = 0.1, mu = 0.1, 
                              seeds_random) |> 
  group_by(t) |> 
  summarize(ninf = n())
# targeted simulation
realisation_targeted <- sim_sir(g, beta = 0.1, mu=0.1,
                                seeds_centrality$inf) |>
  group_by(t) |> 
  summarize(ninf = n())

# USE realisation_random results from Step 3 as the no-quarantine comparison 
# Plot infection curve - Unchanged
ggplot() + 
  geom_line(data = realisation_random,
            aes(x = t, y = ninf,
                col = "Random")) +
  geom_line(data = realisation_targeted,
            aes(x = t, y = ninf,
                col = "Targeted"))


# a) Difference in total infected 
total_random <- sum(realisation_random$ninf)
total_targeted <- sum(realisation_targeted$ninf)
diff_total <- total_targeted - total_random
# b) Difference in time of peak infection 
peak_random <- realisation_random$t[which.max(realisation_random$ninf)]
peak_targeted <- realisation_targeted$t[which.max(realisation_targeted$ninf)]
diff_peak <- peak_random - peak_targeted


```

### Step 5: Minimising Spread

Suppose now that you can convince 5% of people in the network not to spread that information at all.

1.  Choose those 5% randomly in the network. Simulate the SIR model above βc using 1% of the remaining nodes as seeds. Choose those seeds randomly.
2.  Choose those 5% according to their centrality. Simulate the SIR model above βc using 1% of the remaining nodes as seeds. Choose those seeds randomly.
3.  Measure the difference between both cases as you did in step 3.

Workflow: Create two reduced networks - one with 5% random nodes deleted, and one with 5% high-centrality nodes deleted. For both reduced networks, select 1% random seeds and run Original SIR. Again measure the difference in total infected and the difference in time of peak infection (from Step 3).

```{r}
# Code from Tut - Not Changed - CHANGE BETA VALUES

# Random seed vax (change to 5%)
vax_random <- sample(1:vcount(g),
                     vcount(g) * 0.2)
# Random network without vaccinated
gp_random <- delete_vertices(g, vax_random) 
# random spreaders
seeds <- sample(1:vcount(gp_random),
                vcount(gp_random) * 0.01) 
# simulation with random vaccination
realisation_random <- sim_sir(gp_random, 
                              beta = 0.1, mu = 0.1, 
                              seeds) |>
  group_by(t) |> 
  summarize(ninf = n())

# Centrality-based seed vax (change to 5%)
vax_centrality <- closeness(g) |> 
  enframe(name = "inf", 
          value ="closeness") |>
  slice_max(closeness, 
            n = vcount(g) * 0.2) 
# Centrality network without vaccinated
gp_centrality <- delete_vertices(g, vax_centrality$inf) 
# randomn spreaders
seeds <- sample(1:vcount(gp_centrality), 
                vcount(gp_centrality) * 0.01) 

# simulation with centrality-based vaccination
realisation_targeted <- sim_sir(gp_centrality, 
                                beta = 0.5, mu = 0.1, 
                                seeds) |>
  group_by(t) |> 
  summarize(ninf = n())

# NEED TO ADD: Steps a and b 
# a) Difference in total infected 

# b) Difference in time of peak infection 
```

### Step 6: Analysis

1.  Comment on the relationship between the findings in steps 3 and 5 using the same type of centrality for the 1% in step 3 and 5% in step 5.

Workflow: Discuss the results from Step 3 and Step 5. Didn't fully understand this step so asked Claude - response was: "The instruction is saying that whatever centrality metric you chose in step 3 to identify your best 1% seeds (e.g. closeness), you should use that same metric in step 5 to identify the 5% to vaccinate. You are asking whether the nodes that are best at spreading (step 3) are also the most important ones to suppress (step 5). Using different centrality metrics in each step would muddy that comparison."

### Step 7: Predictive Model

1.  Train a model that predicts that time to infection of a node using their degree, centrality, betweeness, page rank or any other predictors you see fit. Use that model to select the seed nodes as those with the smallest time to infection in step 3. Repeat step 5 with this knowledge.

Workflow: Run a simulation (Original SIR function) and random seeds to generate baseline infection times. Extract node features (e.g., degree, centrality, betweenness, page rank, etc.) and merge with the baseline infection times (training data). Fit a model (linear regression?) to predict the time to infection for every node (t) in the network. Repeat the code from Step 3, replacing the centrality-based (1%) seeds with the 1% of nodes with the smallest predicted time (t) and compare it to the random seed (normal simulation). Repeat the code from Step 5, replacing the centrality-based vax targets (5%) with the 5% of nodes that have the smallest predicted time (t) (essentially the nodes identified as spreading the infection earliest) and compare the random vs. model vaccination spread.

Could include Date of posting, With/Without Condom (Anal and Oral), mouth kiss (depending on STI) but they are currently edge-level attributes but model is node-level features (need to aggregate if we want to use them).

#### Part A: Build the Predictive model

##### 1. Generate training data:

```{r}
# Baseline infection times
## TODO: Replace beta = 0.05 with beta_high once Step 2 is complete
seeds_train <- sample(1:vcount(g), vcount(g) * 0.01)
realisation_train <- sim_sir(g, beta = 0.05, mu = 0.1, seeds_train)
```

##### 2. Feature Extraction

```{r}
feat_degree    <- degree(g)             |> enframe(name = "inf", value = "degree")
feat_closeness <- closeness(g)          |> enframe(name = "inf", value = "closeness")
feat_between   <- betweenness(g)        |> enframe(name = "inf", value = "betweenness")
feat_pagerank  <- page_rank(g)$vector   |> enframe(name = "inf", value = "page_rank")
```

##### 3. Merge into one table

```{r}
## Convert inf to character so joins work correctly
feat_degree$inf    <- as.character(feat_degree$inf)
feat_closeness$inf <- as.character(feat_closeness$inf)
feat_between$inf   <- as.character(feat_between$inf)
feat_pagerank$inf  <- as.character(feat_pagerank$inf)

features <- feat_degree |>
  left_join(feat_closeness, by = "inf") |>
  left_join(feat_between,   by = "inf") |>
  left_join(feat_pagerank,  by = "inf")
```

##### 4. Merge features with infection times

```{r}
## Nodes never reached by the simulation are dropped (they have no infection time)
realisation_train$inf <- as.character(realisation_train$inf)
training_data <- features |>
  left_join(realisation_train, by = "inf") |>
  filter(!is.na(t))
```

##### 5. Fit model

```{r}
# model <- lm(t ~ ... , data = )
## Predicts time-to-infection (t) from network structure features
## A lower predicted t = spreads faster = better seed / more important to suppress
model <- lm(t ~ degree + closeness + betweenness + page_rank, data = training_data)
summary(model)  # Check which features are significant predictors
```

##### 6. Predict time to infection for ALL nodes

```{r}
features$pred_t <- predict(model, newdata = features)
```

#### Part B: Re-run Step 3

Instead of picking seeds by centrality (Step 3), we pick the 1% of nodes with the smallest PREDICTED infection time - i.e. predicted to spread fastest

##### 7. Select smallest predicted infection time as seeds

```{r}
seeds_model_s3 <- features |>
  slice_min(pred_t, n = round(vcount(g) * 0.01)) |>
  pull(inf) |> as.integer()
```

##### 8. Run random seed simulation

```{r}
# Random spreaders
seeds_random_s3 <- sample(1:vcount(g), 
                       vcount(g) * 0.01) 

realisation_random_s3 <- sim_sir(g, beta = 0.05, mu = 0.1, 
                              seeds_random_s3) |> 
  # TODO: Replace beta = 0.05 with beta_high once Step 2 is complete
  group_by(t) |> 
  summarize(ninf = n())
```

##### 9. Run the model-based seed simulation

```{r}
# targeted simulation
realisation_model_s3 <- sim_sir(g, beta = 0.05, mu = 0.1,
                                seeds_model_s3) |>
  # TODO: Replace beta = 0.05 with beta_high once Step 2 is complete
  group_by(t) |> 
  summarize(ninf = n())
```

##### 10. Plot comparison

```{r}
# Plot infection curve 
ggplot() + 
  geom_line(data = realisation_random_s3,
            aes(x = t, y = ninf,
                col = "Random")) +
  geom_line(data = realisation_model_s3,
            aes(x = t, y = ninf,
                col = "Model-based")) +
  labs(title = "Step 7: Model-based vs Random Seeds",
       x = "Time", y = "New infections", col = "Strategy")
```

##### 11. Measure differences

```{r}
# a) Difference in total infected
total_random_s3 <- sum(realisation_random_s3$ninf)
total_model_s3  <- sum(realisation_model_s3$ninf)
total_model_s3 - total_random_s3  # positive = model seeds infected more people
```

```{r}
# b) Difference in time of peak infection 
peak_random_s3 <- realisation_random_s3$t[which.max(realisation_random_s3$ninf)]
peak_model_s3  <- realisation_model_s3$t[which.max(realisation_model_s3$ninf)]
peak_random_s3 - peak_model_s3  # positive = model seeds peaked earlier
```

#### Part C: Re-run Step 5 with model-selected vaccination targe

##### 12. Select model-based vaccination targets (5% smallest predicted t)

```{r}
# Random seed vax (change to 5%)
vax_model <- features |>
  slice_min(pred_t, n = round(vcount(g) * 0.05)) |>
  pull(inf) |> as.integer()

```

##### 13. Build reduced network with model-selected nodes removed

```{r}
# Random network without vaccinated
gp_model_vax <- delete_vertices(g, vax_model)
```

##### 14. Random seeds on reduced network

```{r}
seeds_remaining <- sample(1:vcount(gp_model_vax), vcount(gp_model_vax) * 0.01)
realization_model_s5 <- sim_sir(gp_model_vax, beta = 0.05, mu = 0.1, seeds_remaining) |>
  # TODO: Replace beta = 0.05 with beta_high once Step 2 is complete
  group_by(t) |>
  summarize(ninf = n())
```

##### 15. Random vaccination baseline

```{r}
vax_random_s5 <- sample(1:vcount(g), vcount(g) * 0.05)
g_random_vax  <- delete_vertices(g, vax_random_s5)
seeds_random_s5 <- sample(1:vcount(g_random_vax), vcount(g_random_vax) * 0.01)
realization_random_s5 <- sim_sir(g_random_vax, beta = 0.05, mu = 0.1, seeds_random_s5) |>
  # TODO: Replace beta = 0.05 with beta_high once Step 2 is complete
  group_by(t) |>
  summarize(ninf = n())
```

##### 16. Plot comparison

```{r}
ggplot() +
  geom_line(data = realization_random_s5,
            aes(x = t, y = ninf, col = "Random Vaccination")) +
  geom_line(data = realization_model_s5,
            aes(x = t, y = ninf, col = "Model-based Vaccination")) +
  labs(title = "Step 7: Model-based vs Random Vaccination",
       x = "Time", y = "New infections", col = "Strategy")
```

##### 17. Measure differences (same metrics as Step 3)

```{r}
# a) Difference in total infected
total_random_s5 <- sum(realization_random_s5$ninf)
total_model_s5  <- sum(realization_model_s5$ninf)
total_random_s5 - total_model_s5  # positive = model vaccination reduced more infections
```

```{r}
# b) Difference in time of peak infection
peak_random_s5 <- realization_random_s5$t[which.max(realization_random_s5$ninf)]
peak_model_s5  <- realization_model_s5$t[which.max(realization_model_s5$ninf)]
peak_model_s5 - peak_random_s5  # positive = model vaccination delayed the peak
```

## References 

Cinelli, M., Quattrociocchi, W., Galeazzi, A., Valensis, C. M., Brugnoli, E., Schmidt, A. L., Zola, P., Zollo, F. & Scala, A. (2020). The COVID-19 social media infodemic. *Scientific Reports*, *10*, Article 16598. <https://doi.org/10.1038/s41598-020-73510-5>
