---
title: "Social Network Analysis | UC3M"
subtitle: "Final Project: Spreading on a Social Network"
author: 
      - Samantha Riordan   | 100572578 
      - Miriam Lim-Leal    | 100453357
      - Noah Emanuel-Fürup | 100564888
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output: 
  html_document: 
  toc: TRUE
  toc_float: TRUE
  toc_depth: 3
  number_sections: TRUE
---

## Objective: Study the problem of spreading on a social network

## Intro

```{r, include=F}
# Libraries 
library(tidyverse)
library(igraph)
library(ggthemes)
```

**Network**: Sexual contact patterns between Brazilian escorts and sex buyers.

Rocha, L. E. C., Liljeros, F., Holme, P. (2011). Simulated Epidemics in an Empirical Spatiotemporal Network of 50,185 Sexual Contacts. *PLOS Computational Biology, 7*(3), Article e1001109. [https://doi.org/10.1371/journal.pcbi.1001109](#0)

### Network Setup

```{r}
# Load Network 
read <- read.csv("network_data.csv", 
                    sep = ";", 
                 # Skip rows in csv beginning with "#"
                    comment.char = "#", 
                    header = FALSE)

# Select relevant vars and tidy 
network <- read |> 
  # Add prefix to distinguish distinct nodes (bipartite)
  mutate(female_node = paste0("f_", V1), 
         male_node = paste0("m_", V2)) |> 
  select(female_node, male_node) |> 
  distinct() # Drop duplicate pairs

# Create the graph 
g <- graph_from_data_frame(network, directed = FALSE)
g <- simplify(g)

# Check if the network connected
is_connected(g) # FALSE
# Identify largest connected component (LCC)
components_g <- components(g)
# Create new graph extracting nodes in LCC
g <- induced_subgraph(g, which(components_g$membership == which.max(components_g$csize)))

# Assign node type as a vertex attribute
# Escorts = TRUE, Buyers = FALSE (as per igraph bipartite convention <-- CHECK THIS)
V(g)$type <- startsWith(V(g)$name, "f_")

# Check graph properties 
is_bipartite(g) # TRUE

# Check number of nodes and edges after extraction
vcount(g) # Number of nodes 
ecount(g) # Number of edges

# Other attributes
# Separate degree distributions
deg_escorts <- degree(g)[V(g)$type == TRUE]
deg_buyers  <- degree(g)[V(g)$type == FALSE]
data.frame(
  type = c("Escorts", "Buyers"),
  n = c(length(deg_escorts), length(deg_buyers)),
  mean = c(mean(deg_escorts), mean(deg_buyers)),
  median = c(median(deg_escorts), median(deg_buyers)),
  sd = c(sd(deg_escorts), sd(deg_buyers)))
```

**Network Attributes**

The network returns a total of 15,810 nodes and 38,540 links following the removal of duplicate links and extracting the largest connected component. The total number of nodes represented by escorts

The mean degree (M = 4.88, 2dp), median (Mdn = 2) and standard deviation (sd = 10.71, 2dp) were checked, indicating evidence of a positively skewed degree distribution with high variability. Given that highly skewed networks imply a small fraction of nodes with disproportionately high degree, and the subsequent impact this has on epidemic thresholds, this skewness will be further explored in the following steps to ensure appropriate selection of βc relative to characteristics and contextual factors of this network.

## Steps

1.  Find the theoretical epidemic threshold βc for your network for the information to reach a significant number of nodes.

2.  Assuming that randomly-selected 1% initial spreaders, simulate the SIR model for a range below and above that threshold and plot the number of infected people as a function of β.

3.  Choose a β well-above above βc. Using centrality, communities or any other suitable metric, find a better set of 1% of seeds in the network so we get more infected people than the random case. Measure the difference of your choice with the random case as: a) The difference in the total number of infected people b)The difference in the time of the peak of infection (when most infections happen).

4.  Using the same β, design a “quarantine strategy”: during the exponential growth phase of the system (typically \~t = 3 , but check) quarantine of the susceptible population. You can model quarantine by temporally removing these nodes. Release the quarantined nodes 8 time steps later, making them susceptible again. Measure the difference with respect to no quarantine.

5.  Suppose now that you can convince 5% of people in the network not to spread that information at all.

    1.  Choose those 5% randomly in the network. Simulate the SIR model above βc using 1% of the remaining nodes as seeds. Choose those seeds randomly.
    2.  Choose those 5% according to their centrality. Simulate the SIR model above βc using 1% of the remaining nodes as seeds. Choose those seeds randomly.
    3.  Measure the difference between both cases as you did in step 3.

6.  Comment on the relationship between the findings in steps 3 and 5 using the same type of centrality for the 1% in step 3 and 5% in step 5.

7.  Train a model that predicts that time to infection of a node using their degree, centrality, betweeness, page rank or any other predictors you see fit. Use that model to select the seed nodes as those with the smallest time to infection in step 3. Repeat step 5 with this knowledge.

## Workflow

0.  Define Original SIR function (original)
1.  Find the theoretical epidemic threshold (mu, kavg, ksq and betac code)
2.  Calculate for each realisation (β range below and above βc) the number of total infections using the epidemic threshold identified in step 1 (Use Original SIR model), and CHECK plot the average total infected vs β to visualise epidemic threshold.
3.  Set random seed and centrality-based spreaders, run a normal simulation using the random seeds and a targeted simulation using centrality-based seeds (both with Original SIR calls), plot the infection curve and then calculate the difference in total infected and the difference the time of peak infection between the random and centrality-based seeds.
4.  Detect the exponential growth phase (using a visual inspection of the infection curve from Step 3 to confirm \~*t* = 3) and modify the Original SIR function to create the Quarantine SIR function. Repeat Step 3 for the quarantine simulation (random seeds) identifying the total infected and time of peak infection and compare this the random seed results from Step 3 (compared as with quarantine vs. no-quarantine strategies). I think we will need to select and justify what % of the susceptible population becomes quarantined.
5.  Create two reduced networks - one with 5% random nodes deleted, and one with 5% high-centrality nodes deleted. For both reduced networks, select 1% random seeds and run Original SIR. Again measure the difference in total infected and the difference in time of peak infection (from Step 3).
6.  Discuss the results from Step 3 and Step 5. Didn't fully understand this step so asked Claude - response was: "The instruction is saying that whatever centrality metric you chose in step 3 to identify your best 1% seeds (e.g. closeness), you should use that same metric in step 5 to identify the 5% to vaccinate. You are asking whether the nodes that are best at spreading (step 3) are also the most important ones to suppress (step 5). Using different centrality metrics in each step would muddy that comparison."
7.  Run a simulation (Original SIR function) and random seeds to generate baseline infection times. Extract node features (e.g., degree, centrality, betweenness, page rank, etc.) and merge with the baseline infection times (training data). Fit a model (logistic regression?) to predict the time to infection for every node (t) in the network. Repeat the code from Step 3, replacing the centrality-based (1%) seeds with the 1% of nodes with the smallest predicted time (t) and compare it to the random seed (normal simulation). Repeat the code from Step 5, replacing the centrality-based vax targets (5%) with the 5% of nodes that have the smallest predicted time (t) (essentially the nodes identified as spreading the infection earliest) and compare the random vs. model vaccination spread.

## Steps

### Step 0: SIR Function (original / no-quarantine)

Workflow: Define Original SIR function (original)

```{r}
# Code from Tut - Not changed 

sim_sir <- function(g, beta, mu, seeds){
  state <- rep(0, vcount(g)) #initial state of the simulation
  state[seeds] <- 1 #infect the seeds
  t <- 0
  table <- data.frame(t = 0, inf = seeds)
  while(sum(state == 1) > 0){
    t <- t + 1
    ## I -> R
    infected <- which(state == 1) # infected nodes
    # generate a random value for every infected, if it's < mu, let the node recover.
    state[infected] <- ifelse(runif(length(infected)) < mu, 2, 1)
    
    ## S -> I
    infected <- which(state == 1) # nodes still infected
    susceptible <- which(state == 0) # susceptible nodes
    contacts <- as.numeric(unlist(adjacent_vertices(g, infected))) # contacts of infected
    contacts <- contacts[contacts %in% susceptible] # get those who are susceptible
    new_infected <- contacts[runif(length(contacts)) < beta] # infect contacts
    new_infected <- unique(new_infected) # remove repeated infections
    if(length(new_infected) > 0){
      state[new_infected] <- 1
      table <- rbind(table, 
                     data.frame(t, inf = new_infected))
    }
  }
  table
}
```

### Step 1: Epidemic Threshold

Find the theoretical epidemic threshold βc for your network for the information to reach a significant number of nodes.

Workflow: Find the theoretical epidemic threshold (mu, kavg, ksq and betac code).

Note:

```{r}
# Code from Tut - Not changed 
mu <- 0.1
kavg <- mean(degree(g))
ksq  <- mean(degree(g)^2)
betac <- mu*kavg/(ksq - kavg)
betac

# NEED TO ADD: Distribution - needed to justify β 
# Visualise degree distribution
degree_df <- data.frame(k = degree(g))

# Linear scale (shows skew)
ggplot(degree_df, aes(x = k)) +
  geom_histogram(bins = 50) +
  labs(title = "Degree Distribution", x = "Degree", y = "Count")

# Log-log scale (reveals power-law / heavy tail structure)
degree_count <- degree_df |>
  count(k) |>
  filter(k > 0)
ggplot(degree_count, aes(x = log10(k), y = log10(n))) +
  geom_point() +
  labs(title = "Degree Distribution (log-log)",
       x = "log10(Degree)", y = "log10(Count)")

```

### Step 2: Simulate SIR

**Centrality-Based Seeds vs Random**

Assuming that randomly-selected 1% initial spreaders, simulate the SIR model for a range below and above that threshold and plot the number of infected people as a function of β.

Workflow: Calculate for each realisation (β range below and above βc) the number of total infections using the epidemic threshold identified in step 1 (Use Original SIR model), and CHECK plot the average total infected vs β to visualise epidemic threshold.

Currently - the β value only gets one realisation - would be more robust if averaged over multiple runs

```{r}
# Code from Tut - Not changed 

# calculate for each realisation the number of total infections, trying for different values of β and fixed μ. 

# beta range
results <- map_dfr(seq(0, 0.2, 0.01), 
        \(beta){seeds <- sample(1:vcount(g),
                                vcount(g) * 0.01)
        # make a realization for each beta
        realization <- sim_sir(g, beta, mu, seeds) 
        # create dataframe row
        data.frame(beta, ninf = nrow(realization)) 
        })
# Visualise
results |> 
  ggplot(aes(x = beta, y = ninf)) + 
  geom_point() +
  geom_vline(xintercept = betac, 
             linetype = 2)
```

### Step 3: Maximising Infection

Choose a β well-above above βc. Using centrality, communities or any other suitable metric, find a better set of 1% of seeds in the network so we get more infected people than the random case. Measure the difference of your choice with the random case as:

a)  The difference in the total number of infected people
b)  The difference in the time of the peak of infection (when most infections happen).

Workflow: Set random seed and centrality-based spreaders, run a normal simulation using the random seeds and a targeted simulation using centrality-based seeds (both with Original SIR calls), plot the infection curve and then calculate the difference in total infected and the difference the time of peak infection between the random and centrality-based seeds.

```{r}
# Code from Tut - Not changed 

# Seed Selection - Unchanged
# Random spreaders
seeds_random <- sample(1:vcount(g), 
                       vcount(g) * 0.01) 
# centrality-based spreaders
seeds_centrality <- closeness(g) |> 
  enframe(name = "inf", 
          value = "closeness") |> 
  slice_max(closeness, 
            n = vcount(g) * 0.01) 

# Simulations - Unchanged - CHANGE BETA VALUES
# normal simulation
realization_random <- sim_sir(g, beta = 0.1, mu = 0.1, 
                              seeds_random) |> 
  group_by(t) |> 
  summarize(ninf = n())
# targeted simulation
realization_targeted <- sim_sir(g, beta = 0.1, mu = 0.1,
                                seeds_centrality$inf) |>
  group_by(t) |> 
  summarize(ninf = n())

# Plot infection curve - Unchanged
ggplot() + 
  geom_line(data = realization_random,
            aes(x = t, y = ninf,
                col = "Random")) +
  geom_line(data = realization_targeted,
            aes(x = t, y = ninf,
                col = "Targeted"))

# NEED TO ADD: Steps a and b 
# a) Difference in total infected 

# b) Difference in time of peak infection 

```

### Quarantine SIR Simulation

Modification needs to:

- Modify existing SIR function to temporarily remove the nodes (acting as a quarantine) that is a random sample of susceptible nodes at \~ t = 3 (check: based on exponential growth phase - Step 3) and then reintroduce those nodes 8 steps later (e.g., \~ t = 11).
- I think we will need to select and justify what % of the susceptible population becomes quarantined.

```{r}
# Original SIR Code from Tut - Not changed 

sim_sir <- function(g, beta, mu, seeds){
  state <- rep(0, vcount(g)) #initial state of the simulation
  state[seeds] <- 1 #infect the seeds
  t <- 0
  table <- data.frame(t = 0, inf = seeds)
  while(sum(state == 1) > 0){
    t <- t + 1
    ## I -> R
    infected <- which(state == 1) # infected nodes
    # generate a random value for every infected, if it's < mu, let the node recover.
    state[infected] <- ifelse(runif(length(infected)) < mu, 2, 1)
    
    ## S -> I
    infected <- which(state == 1) # nodes still infected
    susceptible <- which(state == 0) # susceptible nodes
    contacts <- as.numeric(unlist(adjacent_vertices(g, infected))) # contacts of infected
    contacts <- contacts[contacts %in% susceptible] # get those who are susceptible
    new_infected <- contacts[runif(length(contacts)) < beta] # infect contacts
    new_infected <- unique(new_infected) # remove repeated infections
    if(length(new_infected) > 0){
      state[new_infected] <- 1
      table <- rbind(table, 
                     data.frame(t, inf = new_infected))
    }
  }
  table
}
```

### Step 4: Quarantine Strategy

Using the same β, design a “quarantine strategy”: during the exponential growth phase of the system (typically \~*t* = 3 , but check) quarantine of the susceptible population. You can model quarantine by temporally removing these nodes. Release the quarantined nodes 8 time steps later, making them susceptible again. Measure the difference with respect to no quarantine.

Workflow: Detect the exponential growth phase (using a visual inspection of the infection curve from Step 3 to confirm \~*t* = 3) and modify the Original SIR function to create the Quarantine SIR function. Repeat Step 3 for the quarantine simulation (random seeds) identifying the total infected and time of peak infection and compare this the random seed results from Step 3 (compared as with-quarantine vs. no-quarantine strategies).

```{r}
# Code from Tut - Not Changed - Modify current code to the quarantine function and "Targeted" plot to be "Quarantine" 

# normal simulation
realization_random <- sim_sir(g, beta = 0.1, mu = 0.1, 
                              seeds_random) |> 
  group_by(t) |> 
  summarize(ninf = n())
# targeted simulation
realization_targeted <- sim_sir(g, beta = 0.1, mu=0.1,
                                seeds_centrality$inf) |>
  group_by(t) |> 
  summarize(ninf = n())

# USE realization_random results from Step 3 as the no-quarantine comparison 
# Plot infection curve - Unchanged
ggplot() + 
  geom_line(data = realization_random,
            aes(x = t, y = ninf,
                col = "Random")) +
  geom_line(data = realization_targeted,
            aes(x = t, y = ninf,
                col = "Targeted"))

# NEED TO ADD: Steps a and b 
# a) Difference in total infected 

# b) Difference in time of peak infection 

```

### Step 5: Minimising Spread

Suppose now that you can convince 5% of people in the network not to spread that information at all.

1.  Choose those 5% randomly in the network. Simulate the SIR model above βc using 1% of the remaining nodes as seeds. Choose those seeds randomly.
2.  Choose those 5% according to their centrality. Simulate the SIR model above βc using 1% of the remaining nodes as seeds. Choose those seeds randomly.
3.  Measure the difference between both cases as you did in step 3.

Workflow: Create two reduced networks - one with 5% random nodes deleted, and one with 5% high-centrality nodes deleted. For both reduced networks, select 1% random seeds and run Original SIR. Again measure the difference in total infected and the difference in time of peak infection (from Step 3).

```{r}
# Code from Tut - Not Changed - CHANGE BETA VALUES

# Random seed vax (change to 5%)
vax_random <- sample(1:vcount(g),
                     vcount(g) * 0.2)
# Random network without vaccinated
gp_random <- delete_vertices(g, vax_random) 
# random spreaders
seeds <- sample(1:vcount(gp_random),
                vcount(gp_random) * 0.01) 
# simulation with random vaccination
realization_random <- sim_sir(gp_random, 
                              beta = 0.1, mu = 0.1, 
                              seeds) |>
  group_by(t) |> 
  summarize(ninf = n())

# Centrality-based seed vax (change to 5%)
vax_centrality <- closeness(g) |> 
  enframe(name = "inf", 
          value ="closeness") |>
  slice_max(closeness, 
            n = vcount(g) * 0.2) 
# Centrality network without vaccinated
gp_centrality <- delete_vertices(g, vax_centrality$inf) 
# randomn spreaders
seeds <- sample(1:vcount(gp_centrality), 
                vcount(gp_centrality) * 0.01) 

# simulation with centrality-based vaccination
realization_targeted <- sim_sir(gp_centrality, 
                                beta = 0.5, mu = 0.1, 
                                seeds) |>
  group_by(t) |> 
  summarize(ninf = n())

# NEED TO ADD: Steps a and b 
# a) Difference in total infected 

# b) Difference in time of peak infection 
```

### Step 6: Analysis

1.  Comment on the relationship between the findings in steps 3 and 5 using the same type of centrality for the 1% in step 3 and 5% in step 5.

Workflow: Discuss the results from Step 3 and Step 5. Didn't fully understand this step so asked Claude - response was: "The instruction is saying that whatever centrality metric you chose in step 3 to identify your best 1% seeds (e.g. closeness), you should use that same metric in step 5 to identify the 5% to vaccinate. You are asking whether the nodes that are best at spreading (step 3) are also the most important ones to suppress (step 5). Using different centrality metrics in each step would muddy that comparison."

### Step 7: Predictive Model

1.  Train a model that predicts that time to infection of a node using their degree, centrality, betweeness, page rank or any other predictors you see fit. Use that model to select the seed nodes as those with the smallest time to infection in step 3. Repeat step 5 with this knowledge.

Workflow: Run a simulation (Original SIR function) and random seeds to generate baseline infection times. Extract node features (e.g., degree, centrality, betweenness, page rank, etc.) and merge with the baseline infection times (training data). Fit a model (linear regression?) to predict the time to infection for every node (t) in the network. Repeat the code from Step 3, replacing the centrality-based (1%) seeds with the 1% of nodes with the smallest predicted time (t) and compare it to the random seed (normal simulation). Repeat the code from Step 5, replacing the centrality-based vax targets (5%) with the 5% of nodes that have the smallest predicted time (t) (essentially the nodes identified as spreading the infection earliest) and compare the random vs. model vaccination spread.

Could include Date of posting, With/Without Condom (Anal and Oral), mouth kiss (depending on STI) but they are currently edge-level attributes but model is node-level features (need to aggregate if we want to use them).

```{r}
# Baseline infection times
seeds <- sample(1:vcount(g), vcount(g) * 0.01)
realisation <- sim_sir(g, beta = 0.1, 
                       mu = 0.1, seeds)

# NEED TO ADD:  
# 1. Feature Extraction 
# 2. Merge into one table 

# 3. Fit model 
# model <- lm(t ~ ... , data = )

# 4. Predict time to infection 
# 5. Select smallest predicted infection time as seeds 
```

#### Re-run Step 3

```{r}
# Code from Tut - Not changed 

# Seed Selection - Unchanged
# Random spreaders
seeds_random <- sample(1:vcount(g), 
                       vcount(g) * 0.01) 
# centrality-based spreaders
seeds_centrality <- closeness(g) |> 
  enframe(name = "inf", 
          value = "closeness") |> 
  slice_max(closeness, 
            n = vcount(g) * 0.01) 

# Simulations - Unchanged
# normal simulation
realization_random <- sim_sir(g, beta = 0.1, mu = 0.1, 
                              seeds_random) |> 
  group_by(t) |> 
  summarize(ninf = n())
# targeted simulation
realization_targeted <- sim_sir(g, beta = 0.1, mu=0.1,
                                seeds_centrality$inf) |>
  group_by(t) |> 
  summarize(ninf = n())

# Plot infection curve - Unchanged
ggplot() + 
  geom_line(data = realization_random,
            aes(x = t, y = ninf,
                col = "Random")) +
  geom_line(data = realization_targeted,
            aes(x = t, y = ninf,
                col = "Targeted"))

# NEED TO ADD: Steps a and b 
# a) Difference in total infected 

# b) Difference in time of peak infection 
```

#### Re-run Step 5

```{r}
# Code from Tut - Not Changed

# Random seed vax (change to 5%)
vax_random <- sample(1:vcount(g),
                     vcount(g) * 0.2)
# Random network without vaccinated
gp_random <- delete_vertices(g, vax_random) 
# random spreaders
seeds <- sample(1:vcount(gp_random),
                vcount(gp_random) * 0.01) 
# simulation with random vaccination
realization_random <- sim_sir(gp_random, 
                              beta = 0.1, mu = 0.1, 
                              seeds) |>
  group_by(t) |> 
  summarize(ninf = n())

# Centrality-based seed vax (change to 5%)
vax_centrality <- closeness(g) |> 
  enframe(name = "inf", 
          value ="closeness") |>
  slice_max(closeness, 
            n = vcount(g) * 0.2) 
# Centrality network without vaccinated
gp_centrality <- delete_vertices(g, vax_centrality$inf) 
# randomn spreaders
seeds <- sample(1:vcount(gp_centrality), 
                vcount(gp_centrality) * 0.01) 

# simulation with centrality-based vaccination
realization_targeted <- sim_sir(gp_centrality, 
                                beta = 0.5, mu = 0.1, 
                                seeds) |>
  group_by(t) |> 
  summarize(ninf = n())

# NEED TO ADD: Steps a and b 
# a) Difference in total infected 

# b) Difference in time of peak infection 
```
